Machine Learning Tools for Open Cluster Characterization with Gaia DR2 Data

### 2021-02-09 paper/paper.tex
paper: Add advisor corrections
Add @cguz corrections and suggestions

Co-Authored-By: Cesar Guzmann <15159632+cguz@users.noreply.github.com>

The study of OCs advanced thanks to the huge and precise dataset
from the Gaia DR2 mission, available since 2018 from
\cite{collaboration2016description} and \cite{gaia2018gaia}.
Although, the Gaia DR2 database misreported the chemical composition
of some OCs, the database has helped to review already known OCs and
to find new ones.

@@

Normally, the characterization process considers astrometic features such as
proper motion in right ascension and declination, or parallax.
In some cases~\cite{oliveira2013fitting}, it also considers photometric features,
which helps to generate the H-R diagram~\cite{hypki2018gaia} of the OC candidate stars.
Those OC candidates should present a sharp profile corresponding to an isochrone curve
derived from a theoretical model. The model considers metallicity, mass and brightness
of the stars involved. For all this purpose, we used a set of tools that require
supervised and parameterized models. It means that we need a previous knowledge of the
cluster, otherwise we must repeat the process iteratively to achieve valid results.

In the present paper, we propose a machine learning model capable of characterizing
OCs inside a stellar region with no previous knowledge about the region.
Our model takes advantage of some features such as proper motion in right ascension
and declination, and parallax to train an Artificial Neural Network.
The ANN clusters all the stars within given region in groups.
One of those groups will be the open cluster that interests us.

@@

This paper is organized as follows. Section~\ref{sec:related_works} presents
some related works and Section~\ref{sec:aims_and_method} describes in more detail
our model. The following Section~\ref{sec:results}, shows the results and compares
them with a method that involves tools from the Virtual Observatory (VO) such as
Clusterix~\cite{balaguer2020clusterix}. Finally, the last section concludes and
outlines future research lines.

@@

In~\cite{castro2020hunting}, they present one of the initial approach to detect OCs.
They search for overdensities in the astrometric space of the galactic disk, and then,
they identify with photometric information possible OCs. Although it looks simple,
it is quite difficult to face the problem that the near field around the OC has two
kind of populations. The first population belongs to the membership stars of the OC
(from tens or hundreds to a few thousands), while the second population has a background
of stars that do not belong to the OC (from tens to hundreds of thousands). Thus,
we summarize our problem to find out which stars belong to the OC. This is a handicap
work because the selection is critical to properly characterize the fundamental
properties of the cluster (total mass, age, chemical composition, between others).

Other works, such as TOPCAT~\cite{taylor2005topcat}, Clusterix 2.0~\cite{balaguer2020clusterix},
Aladin~\cite{bonnarel2000aladin} or VOSA~\cite{bayo2008vosa}, perform as a first step
to analyze the proper motion configuration space of the region of interest.

TOPCAT cannot find open cluster, instead it requires to parameterize some of the
properties of the cluster. This last process requires a previous knowledge of the cluster.
In other words, they perform a supervised and manual selection of groups based on
overdensities in the proper motion configuration space. Clusterix 2.0 is an interactive
web-based tool that can help to the selection of groups. It takes the proper motion
diagram without making any prior assumption about the membership of the candidate star
and determines empirically the frequency functions. It employs normal Gaussian kernel
functions, defined as:

@@

In contrast to~\cite{castro2020hunting}, we lack of a super computer like MareNostrum
for our experiments. For this reason, our aim is to obtain a new novel method that
allows the characterization of open clusters without doing a blind search for clusters
characterization. We advocate of the idea of building an unsupervised and
non-parameterized method. So, it an be suitable for automated processes.

@@

Our aim is to \emph{build an unsupervised clustering model for open cluster characterization}.
The model will be \emph{non-supervised and non-parameterized} in order to fit a wide range
of clusters without the need for fine-tuning a high number of hyperparameters.

\subsection{DOWNLOAD}

We start by selecting a region from the OpenClust catalogue and downloading it from
Gaia DR2 database into a docker database. This allows us to repeat the analysis without
having to download the data every time. The radius of the downloaded region from Gaia
is 1.5 times larger than the one registered in OpenClust. This way we ensure to include
several stars that do not belong to the open cluster.

\subsection{FEATURE SELECTION}
\label{sec:feature_selection}

\begin{figure}[htbp]
  \includegraphics[width=\columnwidth]{../figures/melotte_22/features_melotte_22.png}
  \caption{Pairwise relationships among variables using Melotte 22 data}
  \label{fig:features_melotte_22}
\end{figure}

As mentioned before, we want our model to characterize open clusters by looking at
their dynamic properties. Also, we want to maintain as simple as possible our
clustering model in order to save computing resources.

Proper motion in right ascension and declination seems like a natural choice since,
as we know, stars belonging to the same OC share a common motion vector.

Parallax is another important feature. It lets us know how far stars are from us.
In addition, since all stars within an open cluster were born from the same dust cloud,
they must all have similar parallax.

However, we are not going to use these raw features.
Instead, we are taking a combination of them.

First, we correct proper motion in right ascension and declination by dividing them by
the parallax. That way, we normalize these quantities and help our clustering models
to improve their performance.

The modulus of the proper motion is another computed property that we are considering.
We use it to relate both features and therefore, to force our model to keep them tight.

Figure \ref{fig:features_melotte_22} shows a pairwise relationship among some features
available in the dataset for Melotte 22 data.

\subsection{SOFT CLUSTERING}

Our first approach to find the open cluster is using the K-Means algorithm.
Since we are looking for a single cluster, it seems reasonable to use a clustering
algorithm and set it to find two clusters. One for the desired OC and another which
contains stars that do not belong to the open cluster.
However, this idea is not completely right.

This is due to the fact that OC's stars are surrounded by other stars with possibly similar
properties. Therefore, setting the number of clusters to two is too low to separate them properly.

\begin{figure}
  \centering
  \begin{subfigure}{\columnwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n2_pm_melotte_22.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n5_pm_melotte_22.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n8_pm_melotte_22.png}
    \end{subfigure}
  \end{subfigure}
  \medskip
  \begin{subfigure}{\columnwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n2_parallax_melotte_22.png}
      \caption{N = 2}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n5_parallax_melotte_22.png}
      \caption{N = 5}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n8_parallax_melotte_22.png}
      \caption{N = 8}
    \end{subfigure}
  \end{subfigure}
  \caption{K-Means comparisons with Melotte 22}
  \label{fig:kmeans_comparisons_melotte_22}
\end{figure}

As shown in Figure \ref{fig:kmeans_comparisons_melotte_22}, larger values for the number of
clusters allow us to isolate more accurately the resonance in parallax at $\approx 7.3mas$.
However, we have the disadvantage that more groups are formed.

This effect complicates our task of finding the desired open cluster, since we would like
to get just two groups, one for the OC and another with the remaining stars.
Therefore, we have to find a way to set the right value for the number of clusters
to isolate the searched cluster without creating too many groups.

To solve this issue, we will try to estimate the best number of clusters by using the
\emph{silhouette score} \cite{rousseeuw1987silhouettes}.

@@

Therefore, we would like to find a better model that improves this initial characterization
by reducing the amount of clusters and also that removes outsiders from the OC.

\subsection{DEC MODEL}

Since we do not have a labeled dataset to train a supervised model,
we have no choice but to use an unsupervised self-trained model.

For that reason we have adapted the \emph{Unsupervised Deep Embedding for Clustering Analysis}
to our work.

The model is composed by a \emph{deep autoencoder}, a \emph{deep encoder} and
a \emph{clustering layer}.

Encoders are used to transform the input data into a latent space
using a non-linear mapping function $f_{\theta} : X \rightarrow Z$.

Although, as explained in section \ref{sec:feature_selection},
the number of features we are managing is not too large,
this latent space helps us reduce the number of features
and avoids the \emph{"curse of dimensionality"} \cite{bellman1961curse}.

These encoders are pretrained before fitting the model to generate predictions. Then,
the encoder is used with the aim of transforming input data to the latent space $Z$. Once the
data has been transformed, a K-Means clusterer is used in order to make an initial clustering.

With that initial configuration, the model iterates alternating between computing an auxiliary
target distribution (Soft Assignment) and minimizing the Kullback-Leiber (KL) divergence
\cite{kullback1951information} to it. This unsupervised algorithm allows us to improve the clustering.

In the soft assignment stage,
the \emph{Student's t-distribution} is used as a kernel to measure the similarity
between the embedded points and the cluster centroid.
While in the KL divergence minimization the algorithm iteratively refines clusters by learning
from their high confidence assignments with the help of an auxiliary target distribution.
The model is trained by matching the soft assignment to the target distribution.
The choice of this target distribution is crucial for DEC's performance.
In this work we have taken the target distribution from DEC's original paper \cite{xie2016unsupervised},
which is defined in Equation \ref{eq:student_tdistribution}.

Figure \ref{fig:dec_model_setup} shows the layer setup of our DEC model.
It is simpler than the one tested on the original paper \cite{xie2016unsupervised},
since the number of selected features in our work is smaller than in the original one.
Therefore, using the same configuration would result in a model so
powerful that would incur in overfitting issues unable to make right predictions.

\begin{figure}
  \includegraphics[width=\columnwidth]{../figures/dec_diagram.pdf}
  \caption{DEC model layer setup}
  \label{fig:dec_model_setup}
\end{figure}

\begin{figure}[htbp]
  \centering
  \begin{subfigure}{\columnwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_pm_melotte_22.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_pm_melotte_22.png}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_pm_filtered_melotte_22.png}
    \end{subfigure}
  \end{subfigure}
  \medskip
  \begin{subfigure}{\columnwidth}
    \centering
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_hr_diagram_melotte_22.png}
      \caption{K-Means}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_melotte_22.png}
      \caption{DEC}
    \end{subfigure}
    \hfill
    \begin{subfigure}[t]{0.3\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_filtered_melotte_22.png}
      \caption{DEC (filt.)}
    \end{subfigure}
  \end{subfigure}
  \caption{Evolution of Melotte 22 characterization}
  \label{fig:melotte_22_characterization_evolution}
\end{figure}

Figure \ref{fig:melotte_22_characterization_evolution} shows the evolution
of the characterization of Melotte 22. The first column corresponds to the
characterization obtained by applying K-Means alone to the dataset. The middle
column is the characterization made by the DEC model. And the column on the right
shows the characterization made by the DEC and having filtered the quantiles
lower than 0.10 and higher than 0.90. The top row shows the proper motion
configuration space while the bottom one shows the Hertzsprung–Russell diagrams.

All resources developed for this project are available at the GitHub repository:
\href{https://github.com/cdalvaro/machine-learning-master-thesis}{cdalvaro/machine-learning-master-thesis}

It includes the DEC model developed in Python using the Keras framework
and Jupyter Notebooks \cite{Kluyver2016jupyter} with examples.

\section{CONTRIBUTION}

The developed method is successful since it meets all our established requirements:

\begin{itemize}
  \item Our method gives valid results compatible with current methods as described in Section \ref{sec:results}.
  \item As required from the beginning, it is non-supervised and non-parameterized.
  \item The model shows a great performance compared with current methods.
  \item It is able to deal with large and open regions. As an example,
        it has succeeded characterizing Melotte 25 (Hiades) OC. Also,
        the relative proximity of this cluster, 45 parsec, makes it difficult to
        characterize. Clusterix fails characterizing this cluster. In this case,
        our method proposes at least two more non-catalogued clusters. The analysis
        of these new clusters brings out an interesting stellar dynamics in the studied
        region. It could mean that there was a recent interaction among different molecular clouds.
\end{itemize}

\section{RESULTS}
\label{sec:results}

We now present the results obtained after having analyzed Melotte 25.
This is the largest cluster we have studied and one of the most difficult
to characterize by its distribution and extension.

More results are available at Appendix \ref{sec:appendix}.

@@

\begin{table}[htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{l|c|c|c|c}
        \textbf{Method} & \emph{$\mu_{\alpha}$ $(mas/yr^{-1})$} & \emph{$\mu_{\delta}$ $(mas/yr^{-1})$} & \emph{$\varpi$ $(mas)$} & \emph{\# stars} \\
        \hline
        \textbf{Simbad}\tablefootnote{Results have been taken from the SIMBAD astronomical database\cite{wenger2000simbad}} & 104.92 $\pm$ 0.12 & -28.00 $\pm$ 0.09 & 21.052 $\pm$ 0.065 & - \\
        Clusterix & 106.796 $\pm$ 6.229 & -24.870 $\pm$ 5.417 & 21.210 $\pm$ 1.115 & 109 \\
        K-Means & 79.936 $\pm$ 3.7 & -45.362 $\pm$ 3.8 & 20.901 $\pm$ 0.31 & 374 \\
        DEC & 104.051 $\pm$ 3.2 & -33.424 $\pm$ 2.5 & 22.072 $\pm$ 0.13 & 219 \\
        \textbf{DEC (filt.)} & 105.96 $\pm$ 3.5 & -30.00 $\pm$ 2.4 & 21.74 $\pm$ 0.07 & 175 \\
      \end{tabular}
    }
    \caption{Melotte 25 results.}
    \label{tab:results_melotte_25}
  \end{center}
\end{table}

@@

\section{RESULTS DISCUSSION}

@@

\section{CONCLUSIONS}

From the results shown in Section \ref{sec:results}, we can say that, in general,
our model works fine identifying and characterizing open clusters.
So we have succeeded building a model non-parameterized and non-supervised
for open cluster characterization. However, some improvements should be done
in order to improve its accuracy and precision.

@@

\subsection{FURTHER RESEARCH}

@@

\begin{table}[htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{l|c|c|c|c}
        \textbf{Method} & \emph{$\mu_{\alpha}$ $(mas/yr^{-1})$} & \emph{$\mu_{\delta}$ $(mas/yr^{-1})$} & \emph{$\varpi$ $(mas)$} & \emph{\# stars} \\
        \hline
        \textbf{Simbad} & -4.6579 $\pm$ 0.0075 & 11.1517 $\pm$ 0.0075 & 2.4118 $\pm$ 0.0006 & 1727 \\
        Clusterix & -4.652 $\pm$ 0.523 & 11.203 $\pm$ 0.454 & 2.409 $\pm$ 0.127 & 638 \\
        K-Means & -4.344 $\pm$ 0.14 & 9.507 $\pm$ 0.19 & 2.268 $\pm$ 0.01 & 1542 \\
        DEC & -4.426 $\pm$ 0.17 & 9.952 $\pm$ 0.20 & 2.436 $\pm$ 0.01 & 1532 \\
        \textbf{DEC (filt.)} & -4.502 $\pm$ 0.14 & 10.114 $\pm$ 0.17 & 2.392 $\pm$ 0.004 & 1072 \\
      \end{tabular}
    }
    \caption{NGC 2516 results.}
    \label{tab:app_results_ngc_2516}
  \end{center}
\end{table}

>>>>

The study of OCs advanced thanks to the huge and precise dataset
from the Gaia DR2 mission, available since 2018 from~\cite{collaboration2016description}
and~\cite{gaia2018gaia}. Although, the Gaia DR2 database misreported the chemical
composition of some OCs, the database has helped to review already known OCs and
to find new ones.

@@

Normally, the characterization process considers astrometric features such as
proper motion in right ascension and declination, or parallax.
In some cases~\cite{oliveira2013fitting}, it also considers photometric features,
which helps to generate the H-R diagram~\cite{hypki2018gaia} of the OC candidate stars.
Those OC candidates should present a sharp profile corresponding to an isochrone curve
derived from a theoretical model. The model considers metallicity, mass and brightness
of the stars involved. For all this purpose, we used a set of tools that require
supervised and parameterized models. It means that we need a previous knowledge of the
cluster, otherwise we must repeat the process iteratively to achieve valid results.

In the present paper, we propose a machine learning model capable of characterizing
OCs inside a stellar region with no previous knowledge about the region.
Our model takes advantage of some features (such as proper motion in right ascension
and declination, and parallax) to train an Artificial Neural Network.
The ANN clusters all the stars within a given region in groups.
One of those groups will be the open cluster that interests us.

@@

This paper is organized as follows. Section~\ref{sec:related_works} presents
some related works and Section~\ref{sec:aims_and_method} describes in detail
our model. The following Section~\ref{sec:results}, shows the results and compares
them with a method that involves tools from the Virtual Observatory (VO) such as
Clusterix~\cite{balaguer2020clusterix}. Finally, the last section concludes and
outlines future research lines.

@@

In~\cite{castro2020hunting}, they present one of the initial approaches to detect OCs.
They search for overdensities in the astrometric space of the galactic disk. Then,
they identify with photometric information possible OCs. Although it seems easy,
it is quite hard to face the problem that the near field around the OC has two kinds of
populations. The first population belongs to the members of the OC stars (from tens or
hundreds to a few thousand). The second population has a background of stars that do
not belong to the OC (from tens to hundreds of thousands). Thus, we summarize our problem
in finding out which stars belong to the OC.

Other works (TOPCAT~\cite{taylor2005topcat}, Clusterix 2.0~\cite{balaguer2020clusterix},
Aladin~\cite{bonnarel2000aladin}, or VOSA~\cite{bayo2008vosa}), first analyze the proper
motion configuration space of the region of interest. TOPCAT cannot find an open cluster.
Instead, it requires parameterizing some of the properties of the cluster. This last
process requires previous knowledge of the cluster. In other words, they perform a supervised
and manual selection of groups based on overdensities in the proper motion configuration space.
Clusterix 2.0 is an interactive web-based tool that can help the choice of groups. Take the
proper motion diagram without making any previous assumptions about the membership of the
candidate star. And empirically determine the frequency functions. It employs normal Gaussian
kernel functions, defined as:

@@

In contrast to~\cite{castro2020hunting}, we lack of a super computer like MareNostrum
for our experiments. For this reason, our aim is to obtain a new novel method that
allows the characterization of open clusters without doing a blind search for clusters
characterization. We advocate of the idea of building an unsupervised and
non-parameterized method. So, it can be suitable for automated processes.

@@

In the literature, we have several clustering algorithms, such as K-Means, Mean-Shift
Clustering, DBSCAN~\cite{ester1996density} among other, that can help us to achieve
our aim. Each one behaves better according to the distribution of the objects to
clusterize. However, we need to set a large number of clusters to find the open cluster
we are looking for. It complicates the identification of the OC because most of the
time falls in many outliers.

The unsupervised \emph{Deep Embedding for Clustering Analysis} (\emph{DEC})
model~\cite{xie2016unsupervised} is a refinement K-Means based on ANN. It starts with
K-Means. And then, it trains an autoencoder to reduce the feature space and pass them
through a Clustering Layer which refines the previous selection.

With all this in mind, our method is based on DEC model. We adapted it to manage data
groups based on the dynamic properties of the stars. Thus, we have an unsupervised
clustering model for open cluster characterization. The model fits a wide range of
clusters without the need for fine-tuning a high number of hyperparameters.

\subsection{Selection of the catalogue}

In this work we make use of Gaia DR2 since DR3 has not been released in time for us to
include it. Gaia DR2 is a multidimensional dataset obtained by ESA's Gaia mission
(located at L2, 1.5 million kilometers from Earth) and operational since 2014. The
catalogue has high precision and accuracy astrometric data for more than 1.7 billion
stellar sources, and magnitudes in three photometric filters (G, BP and RP) for more
than 1,300 million sources.

We start by selecting a region from the OpenClust~\cite{dias2002new} catalogue and
we downloaded it from Gaia DR2 database. The radius of the downloaded region from Gaia
is 1.5 times larger than the one registered in OpenClust, ensuring to include several
stars that do not belong to the open cluster.

\subsection{Selection of features}
\label{sec:feature_selection}

For the selection of features, we studied the Melotte 22 dataset. Table~\ref{tab:
sample_features_melotte22_region} describes an example of the region Melotte 22. Each
row represents the features of a star.

\begin{table}[!htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{c|c|c}
        % \textbf{source\_id} &
        \emph{pmra $(mas \cdot yr^{-1})$} & \emph{pmdec $(mas \cdot yr^{-1})$} & \emph{parallax $(mas)$} \\
        \hline
        % 64035217900973056 &
        2.848682 & -3.291204 & 0.399680 \\
        % 64035217900973440 &
        0.894901 & -3.445501 & 0.416639 \\
        % 64035217901215616 &
        7.924372 & -0.241281 & 0.397743 \\
        % 64035351043566208 &
        -4.433802 & -2.584965 & 0.410695 \\
        % 64035385403304960 &
        0.055990 & -1.760018 & 0.413813 \\
      \end{tabular}
    }
    \caption{Features of Melotte 22 region.}
    \label{tab:sample_features_melotte22_region}
  \end{center}
\end{table}

Proper motion in right ascension (column pmra) and declination (column pmdec) seems
like a natural choice since stars belonging to the same OC share a common motion vector.
Parallax (column parallax) is another relevant feature. It lets us know how far stars
are from us. Besides, since all stars within an open cluster were born from the same
dust cloud, they must all have similar parallax.

However, we are not going to use these raw features. Instead, we are taking a combination
of them. Figure~\ref{fig:features_melotte_22} shows a pairwise relationship of our
combination of features from Melotte 22.

\begin{figure*}[!hbt]
  \centering
  \includegraphics[scale=0.5]{../figures/melotte_22/features_melotte_22.png}
  \caption{Pairwise relationships among variables using Melotte 22 data}
  \label{fig:features_melotte_22}
\end{figure*}

First, we correct proper motion in right ascension and declination by dividing them by
the parallax (variables pmra\_corr and pmdec\_corr, respectively). That way, we normalize
these quantities and help our clustering models to improve their performance. The modulus
of the proper motion (variable pmmod in Figure~\ref{fig:features_melotte_22}) is another
computed property that we considered. We use it to relate both features and therefore
force our model to keep them tight.

\subsection{Deep Open Clustering of stars}
\label{sec:deep_open_clustering}

In this section we explain in more detail our model.

In order to define our model, one of the main problems is that we do not have a labeled
dataset to train a supervised model. Thus, we have to deal with an unsupervised self-trained
model. We adapted the Unsupervised \emph{DEC} model~\cite{xie2016unsupervised} to our requirements.

Same as the DEC model, we have two main components:

\begin{itemize}
  \item \textbf{deep autoencoder:} It is trained before passing the data through the
        \emph{clustering layer}. It is composed by encoder layers followed by decoder
        layers. Recent research has shown that this autoencoder provides meaningful and
        well-separated representations on real-world datasets~\cite{vincent2010stacked}
        and~\cite{hinton2006reducing} to the DEC model.
  \item \textbf{clustering layer:} This layer recibes data transformed by the autoencoder
        and it is iteratively trained until a convergence criterion is met.
\end{itemize}

The deep autoencoder is used to transform the input data into a latent space using a
non-linear mapping function \(f_{\theta} : X \rightarrow Z\).

As it was described in Section~\ref{sec:feature_selection}, the number of features we
deal is not too large. This latent space helps us to start in a reduced number of
features and avoids the \emph{"curse of dimensionality"}~\cite{bellman1961curse}.

The autoencoder is pretrained before fitting the model to generate predictions. Then,
the encoder layers of the autoencoder are used with the aim of transforming input data to
the latent space \(Z\). Once the data has been transformed, a K-Means clusterer is used in
order to make an initial clustering. K-Means cluster centers are used as the initial
weights for the clustering layer.

With that initial configuration, the model iterates alternating between computing an auxiliary
target distribution (Soft Assignment) and minimizing the Kullback-Leiber (KL)
divergence~\cite{kullback1951information} to it.
This unsupervised algorithm allows us to improve the clustering.

In the soft assignment stage,
the \emph{Student's t-distribution} is used as a kernel to measure the similarity
between the embedded points and the cluster centroid.
While in the KL divergence minimization the algorithm iteratively refines clusters by learning
from their high confidence assignments with the help of an auxiliary target distribution.
The model is trained by matching the soft assignment to the target distribution.
The choice of this target distribution is crucial for DEC's performance.
In this work we have taken the target distribution from DEC's original paper~\cite{xie2016unsupervised},
which is defined in Equation~\ref{eq:student_tdistribution}.

\begin{figure*}[!hbt]
  \centering
  \includegraphics[scale=0.8]{../figures/dec_diagram.pdf}
  \caption{DEC model layer setup}
  \label{fig:dec_model_setup}
\end{figure*}

Figure~\ref{fig:dec_model_setup} shows the layer setup of our DEC model.
It is simpler than the one tested on the original paper~\cite{xie2016unsupervised},
since the number of selected features in our work is smaller than in the original one.
Therefore, using the same configuration would result in a model so
powerful that would incur in overfitting issues unable to make right predictions.

\begin{figure*}[!htbp]
  \centering
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n2_pm_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n2_parallax_melotte_22.png}
      \caption{N = 2}
  \end{subfigure}
  \medskip
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n5_pm_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n5_parallax_melotte_22.png}
      \caption{N = 5}
  \end{subfigure}
  \medskip
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n8_pm_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/kmeans/kmeans_n8_parallax_melotte_22.png}
      \caption{N = 8}
  \end{subfigure}
  \caption{K-Means with Melotte 22}
  \label{fig:kmeans_comparisons_melotte_22}
\end{figure*}

\paragraph{Discussion.}

Since we are looking for a single cluster, it seems reasonable to use a clustering algorithm
like K-Means to find two clusters. One for the desired OC and another for stars that do not
belong to the OC. However, the idea is not completely right. This is due to the fact that OC's
stars are surrounded by other stars with possibly similar properties. Therefore, setting the
number of clusters to two is too low to separate them properly.

Figure \ref{fig:kmeans_comparisons_melotte_22} shows an example. It describes the results of
applying K-Means with N=2, N=5 and N=8 in Melotte 22. As we can see, large values of the number
of clusters allow us to isolate more accurately the resonance in parallax at \(\approx 7.3mas\).

The disadvantage is that it forms more groups of stars, which complicates the task of finding the
desired OC, since we would like to get just two groups. Therefore, with a K-Means we have to find
a way to set the right value for the number of clusters to isolate the searched cluster without
creating too many groups. To solve this issue, we can try to estimate the best number of clusters
by using the \emph{silhouette score}~\cite{rousseeuw1987silhouettes}.

@@

\section{Results}
\label{sec:results}

%% Comparación con K-Means, DEC y DEC filter
\begin{figure*}[!htbp]
  \centering
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_pm_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/melotte_22/kmeans_hr_diagram_melotte_22.png}
      \caption{K-Means}
  \end{subfigure}
  \medskip
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_pm_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_melotte_22.png}
      \caption{DEC}
  \end{subfigure}
  \medskip
  \begin{subfigure}[b]{0.3\textwidth}
    \centering
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_pm_filtered_melotte_22.png}
      \includegraphics[width=\textwidth]{../figures/melotte_22/dec_hr_diagram_filtered_melotte_22.png}
      \caption{DEC (filt.)}
  \end{subfigure}
  \caption{Evolution of Melotte 22 characterization}
  \label{fig:melotte_22_characterization_evolution}
\end{figure*}

We have tested our model with the Melotte 25 dataset, which is one of the most difficult dataset
to characterize OC by its distribution and extension. In order to support the discussion presented
in Section~\ref{sec:deep_open_clustering}, we compared our model against a K-Means classification
algorithm. In addition, we perform a second execution only in our model where we filtered the
quantiles in the Melotte 22 dataset.

Our model was implemented in Python 3.8 using the Keras 2.2 framework and Jupyter
Notebooks~\cite{Kluyver2016jupyter}. All tests were run on an Apple Mac Pro Late 2013 with a
2.7GHz 12-Core Intel Xeon E5-2697v2, and 64GB RAM 1866MHz DDR3. For the GPU, we use a graphic
card AMD FirePro D700 with 6GB~\footnote{All resources developed for this project are available
at \href{https://github.com/cdalvaro/machine-learning-master-thesis}
{https://github.com/cdalvaro/machine-learning-master-thesis}}.

% More results are available at Appendix \ref{sec:appendix}.

Figure~\ref{fig:melotte_22_characterization_evolution} shows the evolution
of the characterization of Melotte 22. The first column corresponds to the
characterization obtained by applying K-Means alone to the dataset. The middle
column is the characterization made by the DEC model. And the column on the right
shows the characterization made by the DEC and having filtered the quantiles
lower than 0.10 and higher than 0.90. The top row shows the proper motion
configuration space while the bottom one shows the Hertzsprung–Russell diagrams.

%% Comparación con K-Means, DEC y DEC filter

@@

\begin{table}[!htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{l|c|c|c|c}
        \textbf{Method} & \emph{$\mu_{\alpha}$ $(mas \cdot yr^{-1})$} & \emph{$\mu_{\delta}$ $(mas \cdot yr^{-1})$} & \emph{$\varpi$ $(mas)$} & \emph{\# stars} \\
        \hline
        \textbf{Simbad}\tablefootnote{Results have been taken from the SIMBAD astronomical database\cite{wenger2000simbad}} & 104.92 $\pm$ 0.12 & -28.00 $\pm$ 0.09 & 21.052 $\pm$ 0.065 & - \\
        Clusterix & 106.796 $\pm$ 6.229 & -24.870 $\pm$ 5.417 & 21.210 $\pm$ 1.115 & 109 \\
        K-Means & 79.936 $\pm$ 3.7 & -45.362 $\pm$ 3.8 & 20.901 $\pm$ 0.31 & 374 \\
        DEC & 104.051 $\pm$ 3.2 & -33.424 $\pm$ 2.5 & 22.072 $\pm$ 0.13 & 219 \\
        \textbf{DEC (filt.)} & 105.96 $\pm$ 3.5 & -30.00 $\pm$ 2.4 & 21.74 $\pm$ 0.07 & 175 \\
      \end{tabular}
    }
    \caption{Melotte 25 results.}
    \label{tab:results_melotte_25}
  \end{center}
\end{table}

@@

\section{Result discussions}

@@

\section{Conclusions}

From the results shown in Section~\ref{sec:results}, we can say that, in general,
our model works fine identifying and characterizing open clusters.
So we have succeeded building a model non-parameterized and non-supervised
for open cluster characterization. However, some improvements should be done
in order to improve its accuracy and precision.

Another achievement is that we have developed a model that does not require
complex hardware. Instead, it can be run on workstations with a common GPU,
making it accessible to be implemented in many research centers.

@@

Furthermore, our model is able to deal with large and open regions. As an example,
it has succeeded characterizing Melotte 25 (Hiades) OC. Also, the relative proximity
of this cluster, 45 parsec, makes it difficult to characterize. Clusterix fails
characterizing this cluster. In this case, our method proposes at least two more
non-catalogued clusters. The analysis of these new clusters brings out an interesting
stellar dynamics in the studied region. It could mean that there was a recent
interaction among different molecular clouds.

\subsection{Future works}

@@

\begin{table}[htbp]
  \begin{center}
    \resizebox{\columnwidth}{!}{
      \begin{tabular}{l|c|c|c|c}
        \textbf{Method} & \emph{$\mu_{\alpha}$ $(mas \cdot yr^{-1})$} & \emph{$\mu_{\delta}$ $(mas \cdot yr^{-1})$} & \emph{$\varpi$ $(mas)$} & \emph{\# stars} \\
        \hline
        \textbf{Simbad} & -4.6579 $\pm$ 0.0075 & 11.1517 $\pm$ 0.0075 & 2.4118 $\pm$ 0.0006 & 1727 \\
        Clusterix & -4.652 $\pm$ 0.523 & 11.203 $\pm$ 0.454 & 2.409 $\pm$ 0.127 & 638 \\
        K-Means & -4.344 $\pm$ 0.14 & 9.507 $\pm$ 0.19 & 2.268 $\pm$ 0.01 & 1542 \\
        DEC & -4.426 $\pm$ 0.17 & 9.952 $\pm$ 0.20 & 2.436 $\pm$ 0.01 & 1532 \\
        \textbf{DEC (filt.)} & -4.502 $\pm$ 0.14 & 10.114 $\pm$ 0.17 & 2.392 $\pm$ 0.004 & 1072 \\
      \end{tabular}
    }
    \caption{NGC 2516 results.}
    \label{tab:app_results_ngc_2516}
  \end{center}
\end{table}
